---
title: "Gapminder exploration"
author: "Zoe Vernon"
date: "8/31/2018"
header-includes:
   - \usepackage{float}
output: 
  pdf_document:
    number_sections: true
---

<!-- Thanks to Rebecca Barter for sharing this with me-->

# Introduction

Gapminder is an excellent organization aimed at increasing the use and understanding of statistics on a number of global topics. They collect a variety of data from many sources and aim to produce fact-based statistics reflecting the current state of our world. The data we are exploring throughout this analysis consists of population, life expectency and GDP information for many countries through time.

The data can be found from https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv if you would like to download it yourself.

```{r packages, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
```

```{r scripts, echo = FALSE, warning=FALSE, message=FALSE}
source("R/load.R")
source("R/clean.R")
source("R/utils.R")
```

Fortunately, the data was already very clean, so we did not conduct any major modifications to the data.


```{r load-data, echo = FALSE}
# load the original gapminder data
gapminder_orig <- loadGapminderData()

# clean the gapminder data
gapminder <- cleanGapminderData(gapminder_orig)
```


# Visualizing the gapminder data (ggplot2)


We are interested in exploring life expectancy as a function of GDP. Figure \ref{fig:gdp-life} shows a scatterplot of life expectancy versus GDP.

```{r gdp-life, echo = FALSE, fig.width = 6, fig.height = 4, fig.align="center", fig.cap = "Life expectancy versus GDP for all countries in the year 2007", fig.pos = "H"}
# start with a blank ggplot layer. Input the gapminder data frame.
gapminder %>%
  filter(year == 2007) %>%
  ggplot() +
  # add a scatter layer
  geom_point(aes(x = gdp_per_cap, y = life_exp, size = population, 
                 col = continent), alpha = 0.6) +
  # label the outlying data points
  geom_text(aes(x = gdp_per_cap + 500, y = life_exp, label = country), 
            hjust = 0,
            size = 2,
            data = filter(gapminder, 
                          gdp_per_cap > 9000, life_exp < 60, year == 2007)) +
  # add a nice theme
  theme_nice +
  scale_x_log10() +
  xlab("GDP per capita (log10)") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy versus GDP for 2007")
```


It certainly appears as though there is some kind of rapid increase in the low GDP range, which slows to a gradual increase in the high GDP range. Several African countries have surprisingly low life expectency for their GDP.

Next, we explore change in life expectancy over time. Figure \ref{fig:life-time} shows a series of boxplots, one for each year-continent combination. Each data point corresponds to the life expectency of a country for the given year in the given continent.




```{r life-time, echo = FALSE, fig.width = 8, fig.height = 6, fig.align="center", fig.cap = "Life expectancy over time", fig.pos = "H"}
gapminder %>%
  # remove Oceania
  filter(continent != "Oceania") %>%
  # input data frame into ggplot base function
  ggplot() +
  # add a boxplot layer grouped by year
  geom_boxplot(aes(x = year, y = life_exp, group = year), fill = "grey90") +
  # add a nice theme
  theme_nice +
  xlab("Year") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time") +
  # make a separate plot for each continent
  facet_wrap(~continent)
```


We see that the life expectancy increased in Africa from 1950 up until the 1990s but has stayed fairly constant with a median of around 50 years since the 1990s. The Americas, Asia, and Europe on the other hand, have experienced continued growth. 

## Comparing GDP across continents (dplyr)

Table 1 compares GDP per capita across continents.

```{r echo = FALSE, eval = FALSE}
# calculate the avergae GDP per cap for different contents

# the following code is very inneficient (it violates DRY)
# it is also hard to read
mean(gapminder[gapminder$continent == "Africa", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Europe", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Americas", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Asia", ]$gdp_per_cap)
```


```{r gdp_table, echo = FALSE}
# the following is less repetitive, and much more readible!
gdp_table <- gapminder %>%
  # filter to rows from the year 2007
  filter(year == 2007) %>%
  # calculate the average GDP for all continents
  group_by(continent) %>%
  summarise(countries = n(), # add up the number of countries
            mean = mean(gdp_per_cap), # average GDP
            SD = sd(gdp_per_cap)) %>%
  # arrange the rows in order of decreasing average GDP
  arrange(desc(mean))
# print a LaTeX table displaying this data
kable(gdp_table,
      caption = paste("A table displaying the mean and standard deviation",
                      "of GDP per capita in 2007 for each continent"))
```


Clearly Oceania is leading the way in terms of GDP per cap.

Next, we want to ask about raw GDP (i.e. overall GDP for each country, rather than standardized by per capita). Table 2 shows the average total GDP for each continent for 2007.


```{r total_gdp_table, echo = FALSE}
# the following is less repetitive, and much more readible!
total_gdp_table <- gapminder %>%
  # filter to rows from the year 2007
  filter(year == 2007) %>%
  # define a variable equal to the total GDP
  mutate(gdp_total = gdp_per_cap * population / 1000000000) %>%
  # calculate the average GDP for all continents
  group_by(continent) %>%
  summarise(countries = n(), # add up the number of countries
            mean = round(mean(gdp_total)), # average GDP
            SD = round(sd(gdp_total))) %>%
  # arrange the rows in order of decreasing average GDP
  arrange(desc(mean))
# print a LaTeX table displaying this data
kable(total_gdp_table, 
      caption = paste("A table displaying the mean and standard deviation",
                      "of GDP (in billions) in 2007 for each continent"))
```


## Using ```tidyr()``` with the gapminder data 
See http://swcarpentry.github.io/r-novice-gapminder/14-tidyr/ for more ways to use ```tidyr()``` on this data.

The gapminder data that we used for visualization was already in a clean usuable format.  Here we are given a dataset that requires some processing to get in a more useful form.  

```{r, load_wide_data, echo = TRUE, message = FALSE, warning = FALSE}
# load the wide format of the data
gapminder_wide <- loadGapminderData(path = "data/", 
                                    filename = "gapminder_wide.csv")

# look at wide_data comapred to original 
dim(gapminder_wide)
dim(gapminder)
head(gapminder_wide)[, c(1:4, 37:38)]
head(gapminder)
```

We can that the wide version now has a separate column for each year of GDP, life expectancy, and population.  This data becomes much easier to work with and understand if we can make year into a column.  First, we use the ```gather()``` function to create a long version of the data where we only have 4 columns, one each for continent and country, one (obstype_year) that gives the key for the values (obs_values) in the fourth column.

```{r, to_long, echo = TRUE, warning = FALSE, message = FALSE}
gapminder_long <- gapminder_wide %>% 
  gather(key   = obstype_year, # name new column for ID variable
         value = obs_values,   # name column that will contain the observations
         starts_with('pop'),             # only include columns that start  
         starts_with('lifeExp'),         # pop, lifeExp, or gdpPercap
         starts_with('gdpPercap')) 

head(gapminder_long)
tail(gapminder_long)

```


Next, we separate the obstype_year variable into two separate columns, one that contains the year of the observation and another that tells whether the values in obs_values is gdp, life expectancy, or population. 
```{r echo = TRUE, warning = FALSE, message = FALSE}
# separate the obstype_year into two 
gapminder_long <- gapminder_long %>% 
  separate(obstype_year,
           into = c('obs_type','year'),
           sep="_")

head(gapminder_long)
tail(gapminder_long)
```

Finally, we spread the long version of the data to get the original intermediate version.

```{r, to_normal, echo = TRUE, warning = FALSE, message = FALSE}
gapminder_normal <- gapminder_long %>% 
  spread(obs_type, obs_values)

# check that is looks like the original data
head(gapminder_normal)
dim(gapminder_normal)
dim(gapminder_long)
```