---
title: "Gapminder exploration"
author: "Zoe Vernon"
date: "8/31/2018"
header-includes:
   - \usepackage{float}
output: 
  pdf_document:
    number_sections: true
---

<!-- Thanks to Rebecca Barter for sharing this with me-->

# Introduction

Gapminder is an excellent organization aimed at increasing the use and understanding of statistics on a number of global topics. They collect a variety of data from many sources and aim to produce fact-based statistics reflecting the current state of our world. The data we are exploring throughout this analysis consists of population, life expectency and GDP information for many countries through time.

The data can be found from https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv if you would like to download it yourself.

```{r packages, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
```

```{r scripts, echo = FALSE, warning=FALSE, message=FALSE}
source("R/load.R")
source("R/clean.R")
source("R/utils.R")
```

Fortunately, the data was already very clean, so we did not conduct any major modifications to the data.


```{r load-data, echo = FALSE}
# load the original gapminder data
gapminder_orig <- loadGapminderData()

# clean the gapminder data
gapminder <- cleanGapminderData(gapminder_orig)
```


# Visualizing the gapminder data (ggplot2)


1. We are interested in exploring life expectancy as a function of GDP. Create scatterplot of life expectancy versus GDP using ```ggplot()``` where the size of points are based on the population of the country and they are colored by the continent the country resides in. 

```{r gdp-life, echo = FALSE, fig.width = 6, fig.height = 4, fig.align="center", fig.cap = "Life expectancy versus GDP for all countries in the year 2007", fig.pos = "H"}

```

2. Next, we explore change in life expectancy over time. Create a series of boxplots using ```ggplot()```, one for each year-continent combination, where each data point corresponds to the life expectency of a country for the given year in the given continent.

```{r life-time, echo = FALSE, fig.width = 8, fig.height = 6, fig.align="center", fig.cap = "Life expectancy over time", fig.pos = "H"}

```

## Comparing GDP across continents (dplyr)

1. Compute the mean and variance of the GDP for each continent without using ```dplyr()```

```{r echo = FALSE, eval = FALSE}
# calculate the avergae GDP per cap for different contents

# the following code is very inneficient (it violates DRY)
# it is also hard to read
mean(gapminder[gapminder$continent == "Africa", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Europe", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Americas", ]$gdp_per_cap)
mean(gapminder[gapminder$continent == "Asia", ]$gdp_per_cap)
```

2. Perform the same computation using ```group_by()``` and ```summarize()``` and print the table using the ```kable()``` function

```{r gdp_table, echo = FALSE}

```


3. Next, we want to ask about raw GDP (i.e. overall GDP for each country, rather than standardized by per capita). Create a table that shows the average total GDP for each continent in 2007.


```{r total_gdp_table, echo = FALSE}

```


## Using ```tidyr()``` with the gapminder data 

The gapminder data that we used for visualization was already in a clean usuable format.  Here we are given a dataset that requires some processing to get in a more useful form.  Transform the ```gapminder_wide``` dataset so that it is in the same form as the original gapminder dataset.  

```{r, load_wide_data, echo = TRUE, message = FALSE, warning = FALSE}
# load the wide format of the data
gapminder_wide <- loadGapminderData(path = "data/", 
                                    filename = "gapminder_wide.csv")

# look at wide_data compared to original 
dim(gapminder_wide)
dim(gapminder)
head(gapminder_wide)[, c(1:4, 37:38)]
head(gapminder)
```

1. Use the ```gather()``` and ```separate()``` functions to create a long version of the data where we only have five columns: continent, country, the value of an observation, the type of observation (i.e. GDP, life expectancy, or population size), and the year of the observation. 
```{r, to_long, echo = TRUE, warning = FALSE, message = FALSE}

```

2. Finally, use ```spread()``` to convert the long version of the data to get the original intermediate version.

```{r, to_normal, echo = TRUE, warning = FALSE, message = FALSE}

```

