---
title: "Tidyverse examples"
output: html_document
date: "8/24/2018"
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

This is an introduction to the basic functions in tidyverse, which contains multiple libaries that are useful for manipulating and plotting data in R. 

# dyplr
## piping 
Piping is a way to chain function together to avoid redefining variables.  Below we look at the head of the iris data frame using piping.
```{r piping, message = FALSE, warning = FALSE}
# without using piping
head(iris)

# produces same result but with piping
iris %>% head()
```

## filter
Now we look at ```filter()``` which finds rows where the specified condition is true and returns those rows as a data frame. Here we use ```filter()``` to get only the rows in the data frame that are from the versicolor species.  

```{r filter, message = FALSE, warning = FALSE}
# using filter() without piping
head(filter(iris, Species == "versicolor"))

# same thing but using piping instead
iris %>% 
  filter(Species == "versicolor") %>% 
  head()

# check dimension of data frame
iris %>% 
  filter(Species == "versicolor") %>% 
  dim()
```

Instead of just filtering out one species we could look at all observations that are versicolor as well as setosa.  

```{r filter_two, message = FALSE, warning = FALSE}
# save all rows that are veriscolor of setosa 
iris %>% 
  filter(Species %in% c("versicolor", "setosa")) %>% 
  head()

# check dimension of data frame
iris %>% 
  filter(Species %in% c("versicolor", "setosa")) %>% 
  dim()
```


## select
Now we look at ```select()``` which keeps or removes only certain variables (selecting columns of the data frame, whereas before we were filtering by row). Here we use ```select()``` to keep only the columns that contain the species and sepal length information

```{r select, message = FALSE, warning = FALSE}
iris %>% 
  filter(Species == "versicolor") %>% 
  select(Sepal.Length, Species) %>% 
  head()
```

We can also select all the columns except for sepal length by typing

```{r select_multiple, message = FALSE, warning = FALSE}
iris %>% 
  filter(Species == "versicolor") %>% 
  select(-Sepal.Length) %>% 
  head()
```
## mutate
Now we look at ```mutate()``` which creates new variables consisting of functions of existing variables. First we use ```mutate()``` to create a new variable that is the sum of sepal length and sepal width

```{r mutate_sum, message = FALSE, warning = FALSE}
iris %>% 
  filter(Species == "versicolor") %>% 
  mutate(Sepal.Sum = Sepal.Length + Sepal.Width) %>% 
  head()
```

Next we use ```mutate_at()``` to multiply each Sepal.length and Sepal.Width by 2 

```{r mutate_multiple, message = FALSE, warning = FALSE}
iris %>% 
  filter(Species == "versicolor") %>% 
  mutate_at(vars(contains("Sepal")), funs(2 * .)) %>% 
  head()
```

## group_by and summarize
 ```group_by()``` is a function that changes from operating on the entire dataset to operating on it group-by-group, while ```summarize()``` allows us to summarize the group into a single value.  Below we group the dataset by species and take the mean of t

```{r group_by, message = FALSE, warning = FALSE}
iris %>% 
  group_by(Species) %>%  
  summarize(mean.length = mean(Sepal.Length), n = n())
```
# tidyr
tidyr contains functions for changing the shape of the data, allowing you to transition columns into rows and rows into columns.  The two main functions are ```spread()``` and ```gather()```

## gather
```{r gather, message = FALSE, warning = FALSE}
# top
iris %>% 
  mutate(row = row_number()) %>%
  gather(key = "Variable", value = "Value", -Species) %>%
  head()

# bottom 
iris %>% 
  mutate(row = row_number()) %>%
  gather(key = "Variable", value = "Value", -Species) %>%
  tail()
```

## spread
```{r spread, message = FALSE, warning = FALSE}
iris_long <- iris %>% 
  mutate(row = row_number()) %>%
  gather(key = "Variable", value = "Value", -Species, -row)

iris_wide <- iris_long %>% 
  spread(key = Variable, value = Value)

iris_wide %>% head()
```










